---
title: "Analysis of Environmental Factors: Population, State Parks, and Air Quality"
output: 
  flexdashboard::flex_dashboard:
    theme: yeti
runtime: shiny
---

```{r global, include=FALSE}
#load libraries
library(tidyverse)
library(stringr)
library(rvest)
library(polite)
library(sf)
library(maps)
library(viridis)
library(leaflet)
library(htmltools)
library(janitor)
library(httr)
library(shiny)
library(wordcloud)
library(ggrepel)
```

Introduction
=================
#### Background Information
In this project, we explore how environmental data, specifically air quality, is related to population and state parks. We are interested in examining this relationship and its connection to the number and types of parks found within each state as we investigate broader impacts on air quality. We were motivated to engage with this data because of our love of the outdoors, our experiences visiting state and national parks, and the significance of air quality in connection with climate change and pollution. 

The state parks data for this project comes from the America’s Parks website, and air quality and population data were collected from World Population Review. Air quality and population data are from 2024. The air quality variables used in this project are air quality index (AQI), toxin concentration, and unhealthy days. Population variables include total population, population density, and the percent change in population. 

- AQI is calculated based on the four major air pollutants regulated in the Clean Air Act:          ground-level ozone, carbon monoxide, particle pollution, and sulfur dioxide. AQI ranges from 0    to 500, with 0 indicating very clean air and 500 representing hazardous air. AQI values           falling between 0 to 50 are deemed satisfactory, and values over 50 can cause respiratory         symptoms for sensitive populations. 
- Toxin concentration refers to air pollutants and toxins released from industrial sources that     can have detrimental effects on air quality and human health. Toxin concentration is measured     in pounds per square mile (pounds/mile2).
- Unhealthy days signifies the number of days in a year that have unhealthy air quality.
- Population density represents the number of people per square mile (people/mile2).


#### Data Sources:

- [State Parks](https://www.americasparks.com/index.html)
- [Air Quality](https://worldpopulationreview.com/state-rankings/air-quality-by-state)
- [Population](https://worldpopulationreview.com/states)


#### Github:
The code for this project is linked [here](https://github.com/oceminsky/MiniProject2).


```{r, echo = FALSE}
#function to scrape
scrape_parks <- function(url) {
  Sys.sleep(0.5)
  session <- bow(url)
  scrape(session) |>
    html_nodes("#parklink") |>
    html_text()
}

#scrape the websites
ga_parks <- scrape_parks("https://stateparks.com/ga.html")
mn_parks <- scrape_parks("https://stateparks.com/mn.html")
wi_parks <- scrape_parks("https://stateparks.com/wi.html")
ia_parks <- scrape_parks("https://stateparks.com/ia.html")
ca_parks <- scrape_parks("https://stateparks.com/ca.html")
ut_parks <- scrape_parks("https://stateparks.com/ut.html")
wa_parks <- scrape_parks("https://stateparks.com/wa.html")
me_parks <- scrape_parks("https://stateparks.com/me.html")
az_parks <- scrape_parks("https://stateparks.com/az.html")
sd_parks <- scrape_parks("https://stateparks.com/sd.html")
nd_parks <- scrape_parks("https://stateparks.com/nd.html")
ne_parks <- scrape_parks("https://stateparks.com/ne.html")
mi_parks <- scrape_parks("https://stateparks.com/mi.html")
oh_parks <- scrape_parks("https://stateparks.com/oh.html")
la_parks <- scrape_parks("https://stateparks.com/la.html")
ms_parks <- scrape_parks("https://stateparks.com/ms.html")
mo_parks <- scrape_parks("https://stateparks.com/mo.html")
il_parks <- scrape_parks("https://stateparks.com/il.html")
co_parks <- scrape_parks("https://stateparks.com/co.html")
ny_parks <- scrape_parks("https://stateparks.com/ny.html")

#function for tibble creation
state_tibble <- function(state_name, state_parks) {
  tib <- as_tibble(state_parks)
  tib$state <- state_name
  return(tib)
}

#labeling
ga_tib <- state_tibble("Georgia", ga_parks)
mn_tib <- state_tibble("Minnesota", mn_parks)
wi_tib <- state_tibble("Wisconsin", wi_parks) 
ia_tib <- state_tibble("Iowa", ia_parks)
ca_tib <- state_tibble("California", ca_parks)
ut_tib <- state_tibble("Utah", ut_parks)
wa_tib <- state_tibble("Washington", wa_parks)
me_tib <- state_tibble("Maine", me_parks)
az_tib <- state_tibble("Arizona", az_parks)
sd_tib <- state_tibble("South Dakota", sd_parks)
nd_tib <- state_tibble("North Dakota", nd_parks)
ne_tib <- state_tibble("Nebraska", ne_parks)
mi_tib <- state_tibble("Michigan", mi_parks)
oh_tib <- state_tibble("Ohio", oh_parks)
la_tib <- state_tibble("Louisiana", la_parks)
ms_tib <- state_tibble("Mississippi", ms_parks)
mo_tib <- state_tibble("Missouri", mo_parks)
il_tib <- state_tibble("Illinois", il_parks)
co_tib <- state_tibble("Colorado", co_parks)
ny_tib <- state_tibble("New York", ny_parks)

#parks data cleaning
midwest_parks <- mn_tib |>
  bind_rows(wi_tib, ia_tib, sd_tib, nd_tib, ne_tib, mo_tib, il_tib, mi_tib, oh_tib) |>
  rename(park = value)


full_parks <- ga_tib |>
  bind_rows(mn_tib, wi_tib, ia_tib, ca_tib, ut_tib, wa_tib, me_tib, az_tib, sd_tib, nd_tib, ne_tib, co_tib, il_tib, oh_tib, la_tib, mi_tib, mo_tib, ms_tib, ny_tib) |>
  rename(park = value)
```


```{r, include=FALSE}
#check URL for permission
robotstxt::paths_allowed("https://worldpopulationreview.com/state-rankings/air-quality-by-state")

#web scraping session beginning
session <- bow("https://worldpopulationreview.com/state-rankings/air-quality-by-state", force = TRUE)

#scrape the content, extract and convert
air_result <- scrape(session) |>
  html_nodes(css = "table") |>
  html_table(header = TRUE, fill = TRUE)

air_result
```

```{r, include=FALSE}
air_quality <- air_result[[1]] |>
  select(State, `Air Quality Index↓`, `Air Quality Rank (US News 2024)`, `Days with Unhealthy Air Quality (US News 2024)`, `Industrial Toxin Concentration (US News 2024) (pounds/mi²)`) |>
  rename(air_quality_index = `Air Quality Index↓`,
         air_quality_rank = `Air Quality Rank (US News 2024)`,
         unhealt_days = `Days with Unhealthy Air Quality (US News 2024)`,
         toxin_concentration = `Industrial Toxin Concentration (US News 2024) (pounds/mi²)`) |>
  filter(!State %in% c("Alaska", "Hawaii", "District of Columbia")) |>
  mutate(toxin_concentration = parse_number(toxin_concentration))

air_quality
```

```{r, include=FALSE}
#check URL for permission
robotstxt::paths_allowed("https://worldpopulationreview.com/states#most-and-least-populous-states")

#web scraping session beginning
session <- bow("https://worldpopulationreview.com/states#most-and-least-populous-states", force = TRUE)

#scrape the content, extract and convert
pop_result <- scrape(session) |>
  html_nodes(css = "table") |>
  html_table(header = TRUE, fill = TRUE)

pop_result
```

```{r, include=FALSE}
#population dataset
population <- pop_result[[1]] |>
  select(State, `2025 Pop.  ↓`, Change, Density) |>
  rename(population =`2025 Pop.  ↓`,
         change = Change,
         density = Density) |>
  mutate(population = parse_number(population),
         pop_density = parse_number(density))
```



State Parks
==================


Row
---------------------------
```{r, echo=FALSE}
inputPanel(
  selectInput("stateInput", label = "Select State",
              choices = c("Illinois", "Iowa", "Michigan", "Minnesota", "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin"), 
              selected = "Minnesota"), #state dropdown
  
  selectInput("parkInput", label = "Select Type of Park",
              choices = c("National Park", "National Forest", "State Park", "State Forest", "National Wildlife Refuge"), 
              selected = "State Park") #park types dropdown
)
```


```{r, echo = FALSE}
#creating wordcloud based on total number of parks
num_parks <- full_parks |>
  count(state)

wordcloud(
  words = num_parks$state, 
  freq = num_parks$n, 
  max.words = 50, 
  random.order = FALSE, 
  rot.per = 0.35,
  scale = c(5, 0.25),
  colors = brewer.pal(6, "Dark2"))
```

In this Wordcloud, the size of the words correlates with how many state parks each state has. 

Row
------------------------
```{r, echo=FALSE}

renderPlot({
  state_park_types <- midwest_parks |>
  filter(state == input$stateInput) |>
  mutate(type_park = str_extract(park, "State Forest|State Park|National Park|National Forest|National Wildlife Refuge"), #separating into different types of parks variable
         type_park = replace_na(type_park, "Other")) |>
  count(type_park)
  ggplot(state_park_types, aes(x = type_park, y = n, fill = type_park)) +
    geom_bar(stat = "identity") +
    theme_classic() +
    scale_fill_viridis(discrete = TRUE, option = "G") +
    labs(title = paste("Number of Parks in", input$stateInput),
         x = "Type of Park",
         y = "Number of Parks",
         fill = "Type of Park")
})
```

```{r, echo=FALSE}

renderPlot({
  state_park_types <- midwest_parks |>
  mutate(type_park = str_extract(park, "State Forest|State Park|National Park|National Forest|National Wildlife Refuge"), 
         type_park = replace_na(type_park, "Other")) |>
  filter(type_park == input$parkInput) |>
  group_by(state) |>
  count(type_park)
  ggplot(state_park_types, aes(x = state, y = n, fill = type_park)) +
    geom_bar(stat = "identity", fill = "forestgreen", color = "black") +
    theme_classic() +
    labs(title = paste(input$parkInput, "Counts in Each State of the Midwest"),
         x = "State",
         y = "Number of Parks",
         fill = "Type of Park")
  
})

```



Air Quality
=================


```{r, echo=FALSE}
state_sf <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson") |>
  rename(State = name)


state_air <- right_join(state_sf, air_quality, by ="State")


bins <-  c(35, 40, 45, 50, Inf)
pal <- colorBin("Greens", domain = state_sf$density, bins = bins)

#Hover labels
state_air <- state_air |>
  mutate(State = str_to_title(State)) |>
  mutate(labels = str_c(State, ": ", round(air_quality_index,2), " AQI (0 = great)"))

labels <- lapply(state_air$labels, HTML)

#creation of interactive plot
leaflet(state_air) |>
  setView(-96, 37.8, 3.5) |> #set initial view
  addTiles() |>
  addPolygons( #polygons for each state
    dashArray = "",
    fillColor = ~colorNumeric("Greens", domain = state_air$air_quality_index)(state_air$air_quality_index),
    weight = 3,
    opacity = 2,
    color = "white",
    fillOpacity = 0.8,
    highlightOptions = highlightOptions( #add highlight while hovering
      weight = 4,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions( #label appearance
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) |>
  addLegend(pal = pal,
            values = ~air_quality_index, 
            opacity = 0.7, 
            title = NULL, 
            position = "bottomright")
```


Air Quality and Population
=================


```{r, echo=FALSE, message=FALSE}
pop_air <- air_quality |>
  left_join(population, by = "State")
```


```{r, echo=FALSE, message=FALSE}
ui <- fluidPage(
  titlePanel("Population and Environmental Impact"), #title
  inputPanel( #drop down menu
  selectInput("impactInput", label = "Select Impact Variable:",
    choices = c("Toxin Concentration (lbs//mi²)" = "toxin_concentration", #"choice name" = "actual variable name"
                "Unhealthy Days in 2024" = "unhealt_days",
                "Population (2025)" = "population",
                "Population Density" = "pop_density"))),
  leafletOutput("ourmap")) #output for interactive map
```

```{r, echo=FALSE, message=FALSE}
bins <- 5 #number of bins for color scheme

state_sf <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson") |> rename(State = name) #states dataset and rename to match our datasets

server <- function(input, output, session) {
  state_air_pop <- right_join(state_sf, pop_air, by ="State") #join by state
  
  #what map should render depending on selected variable
  output$ourmap <- renderLeaflet({
    impact_variable <- input$impactInput #variable from dropdown
    label_text <- ifelse(impact_variable == "toxin_concentration", "Toxin Concentration",
                         ifelse(impact_variable == "unhealt_days", "Unhealthy Days in 2025",
                                ifelse(impact_variable == "population", "Population in 2025",
                                       ifelse(impact_variable == "pop_density", "Population Density", ""))))
    pal <- colorBin("RdYlBl", domain = state_air_pop[[impact_variable]], bins = bins) 
    labels <- str_c(state_air_pop$State, "'s ", label_text, ": ", round(state_air_pop[[impact_variable]],2))


    
    #creation of interactive plot
leaflet(state_air_pop) |>
  setView(-96, 37.8, 3.5) |> #set initial view
  addTiles() |>
  addPolygons( #polygons for each state
    dashArray = "",
    fillColor = ~colorNumeric("Blues", domain = state_air_pop[[impact_variable]])(state_air_pop[[impact_variable]]),
    weight = 3,
    opacity = 0.8,
    color = "white",
    fillOpacity = 0.8,
    highlightOptions = highlightOptions( #add highlight while hovering
      weight = 4,
      color = "black",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = ~lapply(labels, HTML),
    labelOptions = labelOptions( #label appearance
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) |>
  addLegend(pal = pal,
            values = state_air_pop[[impact_variable]], 
            opacity = 0.7, 
            title = input$impactInput, 
            position = "bottomright")})
}

#launch shiny app with ui and server
shinyApp(ui, server)
```




State Parks, Air Quality, and Population
===========================

Inputs {.sidebar}
----------------------

```{r, echo=FALSE}
inputPanel(
  selectInput("airquality", label = "Air Quality Variable",
              choices = c("Air Quality Index" = "air_quality_index", "Unhealthy Days in 2024" = "unhealt_days", "Toxin Concentration (lbs//mi²)" = "toxin_concentration")), #new names for drop down
  
  selectInput("population", label = "Population Variable",
              choices = c("Population (2025)" = "population", "Population Density" = "pop_density", "% Change in Population" = "change_numeric")), #new names for drop down
  checkboxInput("check_line", "Include Trend Line?", value = FALSE)
)

```


Row
----------------------

```{r, echo=FALSE}

renderPlot({
park_air <- full_parks |>
    group_by(state) |>
    summarize(num_parks = n()) |>
    left_join(air_quality, by = c("state" = "State")) #join with air quality
labels <- c("air_quality_index" = "Air Quality Index", "unhealt_days" = "Unhealthy Days in 2024", "toxin_concentration" = "Toxin Concentration (lbs//mi²)") #creating labels for the plot
selected_label <- labels[input$airquality]
if(input$check_line)
  ggplot(park_air, aes(x = num_parks, y = .data[[input$airquality]])) +
    geom_point(color = "forestgreen") +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_text_repel(aes(label = state)) + #labeling by state
    labs(title = paste(selected_label, "and Number of Parks per State"),
         x = "Number of Parks",
         y = selected_label)
else if (!input$check_line)
  ggplot(park_air, aes(x = num_parks, y = .data[[input$airquality]])) +
    geom_point(color = "forestgreen") +
    geom_text_repel(aes(label = state)) + #labeling by state
    labs(title = paste(selected_label, "and Number of Parks per State"),
         x = "Number of Parks",
         y = selected_label)
})
```

```{r, echo=FALSE}
renderPlot({
parks_pop <- full_parks |>
    group_by(state) |>
    summarize(num_parks = n()) |>
    left_join(population, by = c("state" = "State")) |> #join with population
    mutate(change_numeric = parse_number(change))
    labels <- c("population" = "Population (2025)", "pop_density" = "Population Density", "change_numeric" = "% Change in Population") #creating labels for the plot
selected_label <- labels[input$population]
if(input$check_line)
  ggplot(parks_pop, aes(x = num_parks, y = .data[[input$population]])) +
    geom_point(color = "forestgreen") +
    geom_text_repel(aes(label = state)) + #labeling by state
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = paste(selected_label, "and Number of Parks per State"),
         x = "Number of Parks",
         y = selected_label)
else if(!input$check_line)
  ggplot(parks_pop, aes(x = num_parks, y = .data[[input$population]])) +
    geom_point(color = "forestgreen") +
    geom_text_repel(aes(label = state)) + #labeling by state
    labs(title = paste(selected_label, "and Number of Parks per State"),
         x = "Number of Parks",
         y = selected_label)
})
```



Conclusion
==================

#### Findings

Overall, this project aims to explore the complex relationship between state parks, air quality, and population across the continental United States. Our analysis focused on these three distinct parts to ensure we include the various factors of environmental, conservation, and population factors. Our goal was to uncover trends through various plots that would help give us a broader perspective of the impacts that may contribute to air quality.  

Our analysis uncovered some notable trends regarding state parks, air quality, toxins, and population density across the continental 48 states. California and New York lead the nation in the number of state parks, emphasizing that larger areas with an emphasis on conservation tend to have more parks. In the Midwest, there exists a large distribution with Michigan and Minnesota containing a larger amount of parks compared to the lower numbers from Ohio and Indiana.

In the context of air quality, Utah has the worst air quality with possible causes being related to higher rates of toxin concentration and unhealthy days. On the other hand, Washington stands out as having the best air quality, which may reflect the lower rates of toxin concentration, population parameters, and unhealthy days. There appears to be a slight pattern in air quality index being affected by toxin concentration and unhealthy days, more so than the population parameters. Another note is there appears to be a pattern in regions with the south western part of the country appearing to have worse air quality index rates, this could be due to various other factors such as California's frequent wildfire threats. 
In the context of population and population density, the most populous states include Califonia, Texas, Florida, and New York. While population density in states like New Jersey, Rhode Island, and Massachusetts are quite high. Population in the context of air quality does not appear to have a significant correlation from a bird’s eye view, however, densely populated states typically face more public health challenges and environmental factors. 

Our comparison revealed a few ideas that may need to be explored further through statistical analysis to see if there are concrete correlations between these hypotheses from the plots seen in this analysis. Specifically, we observed states with larger populations tend to have more state parks. As well as states who have a bad air quality index tend to have higher rates of toxin concentration and unhealthy days. 

In conclusion, our findings reveal relationships between state parks, population, and environmental concerns. Regional factors may have an affect on air quality and state parks. As a result, it is important for policymakers to balance conservation efforts to enhance environmental concerns. Further statistical analysis could provide a clearer understanding of these perceived trends to help guide decisions for the future to inform policymakers as well as the community. 



#### References:

*Air Quality by State 2025*. Air Quality by state 2025. (2025). https://worldpopulationreview.com/state-rankings/air-quality-by-state 

Cheng, J., Schloerke, B., Karambelkar, B., Xie, Y., & Posit. (n.d.). Using Leaflet with Shiny. RStudio. Retrieved May 12, 2025, from https://rstudio.github.io/leaflet/articles/shiny.html

Walker, T. (2025). *State parks*. America’s Parks: State Parks, National Parks, along with all of the other parks - AmericasParks.com. https://www.americasparks.com/index.html 

*US States - Ranking by Population 2025*. US states - ranking by population 2025. (2025). https://worldpopulationreview.com/states 
